Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-22 paper/main.tex
checking in minor edits

Before presenting our results, it is worth considering three possible
explanations of why a participant might answer a given question correctly or
incorrectly. One possibility is that the participant simply \textit{guessed}
the answer. A second is that they selected an answer by mistake, despite
``knowing'' the correct answer. In both of these scenarios, the participant's
knowledge about the question's content should be uninformative about their
observed response. A third possibility is that the participant's response
reflects their \textit{actual} knowledge about the question's content. In this
case, we \textit{might} expect to see a positive relationship between the
participant's knowledge and their likelihood of answering the question
correctly. However, in order to see this positive relationship, the
participant's knowledge must be structured in a way that is reflected (at least
partially) by the embedding space. In other words, if the participant's
performance reflects their true knowledge, but our text embedding space does
not sufficiently capture the structure of that knowledge, then the knowledge
estimates we generate will not be predictive of the participant's performance.
In the extreme, if the embedding space is completely unstructured with respect
to the content of the quiz questions, then we would expect to see the negative
relationship between estimated knowledge and performance that we described
above.

@@

Finally, we used this approach to estimate participants' knowledge about
held-out questions from one lecture using their performance on questions from
the \textit{other} lecture. Here we again observed a similar pattern of
results, though with some notable differences. On Quiz~1, we found that
participants' abilities to correctly answer questions about \textit{Four
Fundamental Forces} could be predicted from their responses to questions about
\textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\
\textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and similarly, that their
ability to correctly answer \textit{Birth of Stars}-related questions could be
predicted from their responses to \textit{Four Fundamental Forces}-related
questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\ \textnormal{CI} = [5.656,
6.843],\ p = 0.043$). Given the results from our analyses that included all
questions and within-lecture predictions, we were surprised to find that the
knowledge estimates could reliably (if weakly) predict participants'
performance across content from different lectures. It is possible that this
result reflects a combination of random guessing prior to training (leading to
a weak effect size), alongside some coarse-scale structured knowledge that
participants had about the content prior to watching either lecture. When we
repeated this analysis using questions from Quiz~2, we found participants'
responses to \textit{Four Fundamental Forces}-related questions did
\textit{not} reliably predict their success on \textit{Birth of Stars}-related
questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027,
3.600],\ p = 0.125$), nor did their responses to \textit{Birth of
Stars}-related questions reliably predict their success on \textit{Four
Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\
95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). These ``prediction
failures'' appear to come from the fact that any signal derived from
participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is swamped by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we carried out these across-lecture knowledge predictions using
questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.017$) using responses to questions about the other lecture's content.
Across all three versions of these analyses, our results suggest that (by and
large) our knowledge estimates can reliably predict participants' abilities to
answer individual quiz questions, distinguish between questions about similar
content, and generalize across content areas, provided that participants' quiz
responses reflect a minimum level of ``real'' knowledge about both content on
which these predictions are based and that for which they are made. Our results
also indicate some important limitations of our approach: if participants' quiz
performance does not reflect what they know (e.g., when they ``guess''), or if
their knowledge is not structured in a way that is reflected by the embedding
space, then our knowledge estimates will not be predictive of their
performance.

>>>>

Before presenting our results, it is worth considering three possible
explanations of why a participant might answer a given question correctly or
incorrectly. One possibility is that the participant simply \textit{guessed}
the answer. A second is that they selected the incorrect answer by mistake,
despite ``knowing'' the correct answer (or vice versa). In both of these
scenarios, the participant's knowledge about the question's content should be
uninformative about their observed response. A third possibility is that the
participant's response reflects their \textit{actual} knowledge about the
question's content. In this case, we \textit{might} expect to see a positive
relationship between the participant's knowledge and their likelihood of
answering the question correctly. However, in order to see this positive
relationship, the participant's knowledge must be structured in a way that is
reflected (at least partially) by the embedding space. In other words, if the
participant's performance reflects their true knowledge, but our text embedding
space does not sufficiently capture the structure of that knowledge, then the
knowledge estimates we generate will not be predictive of the participant's
performance. In the extreme, if the embedding space is completely unstructured
with respect to the content of the quiz questions, then we would expect to see
the negative relationship between estimated knowledge and performance that we
described above.

@@

Finally, we used this approach to estimate participants' knowledge about
held-out questions from one lecture using their performance on questions from
the \textit{other} lecture. Here we again observed a similar pattern of
results, though with some notable differences. On Quiz~1, we found that
participants' abilities to correctly answer questions about \textit{Four
Fundamental Forces} could be predicted from their responses to questions about
\textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\
\textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and similarly, that their
ability to correctly answer \textit{Birth of Stars}-related questions could be
predicted from their responses to \textit{Four Fundamental Forces}-related
questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\ \textnormal{CI} = [5.656,
6.843],\ p = 0.043$). Given the results from our analyses that included all
questions and within-lecture predictions, we were surprised to find that the
knowledge estimates could reliably (if weakly) predict participants'
performance across content from different lectures. It is possible that this
result reflects a combination of random guessing prior to training (leading to
a weak effect size), alongside some coarse-scale structured knowledge that
participants had about the content prior to watching either lecture. When we
repeated this analysis using questions from Quiz~2, we found participants'
responses to \textit{Four Fundamental Forces}-related questions did
\textit{not} reliably predict their success on \textit{Birth of Stars}-related
questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027,
3.600],\ p = 0.125$), nor did their responses to \textit{Birth of
Stars}-related questions reliably predict their success on \textit{Four
Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\
95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). These ``prediction
failures'' appear to come from the fact that any signal derived from
participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is overwhelmed by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we carried out these across-lecture knowledge predictions using
questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.017$) using responses to questions about the other lecture's content.
Across all three versions of these analyses, our results suggest that (by and
large) our knowledge estimates can reliably predict participants' abilities to
answer individual quiz questions, distinguish between questions about similar
content, and generalize across content areas, provided that participants' quiz
responses reflect a minimum level of ``real'' knowledge about both content on
which these predictions are based and that for which they are made.
